Drought frequency predicts life history strategies in \emph{Heliophila}

### 2018-12-11 manuscript/manuscript.tex
update citation style

\abstract{
Explaining variation in life history strategies is a long-standing goal
of evolutionary biology. For plants, annual and perennial life histories
are thought to reflect adaptation to environments that differ in the
frequency of environmental stress such as drought. Here we test this
hypothesis in \textit{Heliophila} (Brassicaceae), a diverse genus of
flowering plants native to Africa, by integrating 34 years of
satellite-based drought measurements with 2192 herbaria occurrence
records. Consistent with predictions from classic life history theory,
we find that perennial \textit{Heliophila} species occur in environments
where droughts are significantly less frequent compared to annuals.
These associations are predictive while controlling for phylogeny,
lending support to the hypothesis that drought related natural selection
has influenced the distributions of these strategies. Additionally, the
difference in drought frequency between annual and perennial species
distributions is greatest during the summer and fall, which also appears
to be when annuals are in the seed phase of their life cycle based on
collection dates of annual species. Together, these finding provide
empirical support for classic hypotheses about the drivers of life
history strategy in plants - that perennials out compete annuals in
environments with less frequent drought and that annuals are adapted to
environments with more frequent drought by escaping drought prone
seasons as seeds. 
}

@@

Understanding the causes and consequences of life history variation is a
longstanding goal of ecology and evolutionary biology (Cole, 1954). In
plants, life histories are especially diverse, with herbaceous species
that complete their life cycle in a number of weeks to trees that live
for thousands of years (Brown, 1996). Along this continuum an important
division exists, distinguishing annuals which complete their seed to
seed life cycle within a single calendar year from perennials which can
persist over multiple years. Annual plants flower once, set seed,
senesce, and then die, spending at least some portion of the year as a
seed, where they are relatively protected from environmental stress. In
contrast, perennial plants can continue vegetative growth after
reproduction and must survive conditions experienced during all seasons.
These represent fundamentally different life history strategies, but the
ecological factors that explain their evolution and distributions remain
empirically uresolved (Friedman and Rubin, 2015).\\
Classical theory predict shorter life spans in environments where adult
mortality is high (Charnov and Schaffer, 1973; Franco and Silvertown,
1996; Stearns, 1992). In plants, this has been extended to the
hypothesis that annuality is adaptive when it allows plants to escape
drought (Schaffer and Gadgil, 1975). Lack of water is perhaps the
greatest threat to survival during vegetative or reproductive growth and
annuals can remain dormant (and protected as a seed) during drought.
Thus, environments with greater seasonal drought frequency may select
for annual life histories that complete reproduction prior to drought
prone seasons. Conversely, environments with less frequent drought may
select for perennial species, which may benefit from multiple bouts of
reproduction and competitive advantage by preventing recruitment of
annual species (Corbin and D'Antonio, 2004). These predictions have been
supported by the observation of annuals in arid environments in
\emph{Oryza perennis} (Morishima et al., 1984) and \emph{Oenothera}
(Evans et al., 2005). Additionally, annual and perennial species of
\emph{Nemesia} were qualitatively associated with winter rather and
summer rainfall environments respectively (Datson et al., 2008) and
annual species of \emph{Scorzoneroides} were associated with
environments classified as unpredictable (Cruz-Mazo et al., 2009).
However, whether the history frequency of drought events indeed predicts
the distributions annual or perennial life history strategies has yet to
be tested.\\
Here we combine a long-term global dataset of satellite detected drought
events with metadata from natural history collections to test these
classic hypotheses within the African endemic mustard genus,
\emph{Heliophila} L. (Brassicaceae). If annuality is an adaptive
strategy allowing plants to escape drought prone seasons, then drought
frequency should predict the distribution of life history strategies
across landscapes, and annual species should be more commonly associated
with drought prone regions than perennial species. Furthermore, if
annual species have adapted to escape drought prone seasons,
observations of growing annual species (i.e.~occurring in forms other
than seed) should be rare during drought prone seasons. Phylogenetic
relatedness can influence tests of associations between species' traits
and their environments (Barrett et al., 1996; Felsenstein, 1985), and
therefore we assessed the relationship between life history distribution
and drought frequency in a phylogenetic context.

@@

Remotely sensed data is a powerful tool for characterizing seasonal
patterns in drought because it is less limited in spatial and temporal
scope and resolution than weather stations or field observations
(AghaKouchak et al., 2015). To quantify the frequency of drought during
different seasons across landscapes, we used the remotely sensed
Vegetative Health Index (VHI), which measures landscape scale reductions
in plant cover and temperature conditions characteristic of drought
(Kogan, 2001). Generated from data collected by NOAA AVHRR satellites
since 1981, the VHI combines Normalized Difference Vegetation Index
(NDVI) derived measures of vegetative stress (Vegetative Condition Index
- VCI) with temperature stress indicated by anomalies in thermal spectra
(Temperature Condition Index - TCI). The VHI of year \(y\) during week
\(w\) of \([1,52]\) at pixel \(i\) is derived from the following
equations, where \(n\) is the number of years observed.

@@

Thus, VHI measurements are standardized according to conditions
historically observed at each locations. These measurements have been
validated and generally used for evaluating drought risk and predicting
crop yields in agriculture (e.g., Rojas et al., 2011; Kogan et al.,
2016). But they also present a new tool to study seasonal patterns in
the frequency of drought across environments and to test hypotheses
about the effect of drought on ecological and evolutionary processes
(Kerr and Ostrovsky, 2003). As such, the VHI has been applied recently
to study drought related ecology of natural species and proven useful
for predicting intraspecific variation in drought tolerance traits and
genes (Dittberner et al., 2018; Mojica et al., 2016; J. Monroe, Powell,
et al., 2018). Here, we accessed VHI data at \(16km^2\) resolution from
1981 to 2015
(\url{https://www.star.nesdis.noaa.gov/smcd/emb/vci/VH/vh_ftp.php}) to
characterize the seasonal drought frequencies experienced by annual and
perennial \emph{Heliophila} species.

@@

\emph{Heliophila} is a genus of flowering plants endemic to the southern
portion of Africa including the Cape Floristic and Succulent Karoo
Regions. These are among the most botanically diverse environments on
Earth and the \emph{Heliophila} species occurring there are considered
to make up the most diverse genus of the family Brassicaceae (Mandáková
et al., 2012; Mummenhoff et al., 2005). This genus includes both
perennial and annual species and this change in life history strategy
has likely arisen multiple independent times (Appel and Al-Shehbaz,
1997; Mummenhoff et al., 2005). Furthermore, the fine scale climatic
heterogeneity of Southern Africa is ideal for studying the distribution
of traits in relation to environmental parameters (Sayre et al., 2013).
We used life histories reported by Mummenhoff et al. (2005), grouping
species with annual or perennial life histories. Perenniality was
defined based any form of perennial life history (e.g., herbs, shrubs,
mixed, etc). Because the nature of species reported with mixed traits
were unknown (i.e.~plasticity vs.~genetic variation), we classified
these species here as perennial since they can maintain vegetative
growth after reproduction at least to some capacity.

@@

Botanists have collected and maintained over 350 million botanical
specimens worldwide over the past 300 years. Herbarium specimens and
their associated metadata have been used since the 1960s to study
species' geographical distributions (reviewed by Willis et al. (2017)
and Lang et al. (2018)). And as they become digitized (Soltis, 2017),
these collections have been used to study relationships between trait
distributions, geography, and climate (Davis et al., 2015; Stropp et
al., 2016; Václavı'k et al., 2017; Wolf et al., 2016). To characterize
the distributions of annual and perennial \emph{Heliophila} species, all
records for the genus \emph{Heliophila} were downloaded from the Global
Biodiversity Information Facility (gbif.org) on July 21, 2018 (GBIF,
2018).

@@

An alignment of ITS I and II sequences for \emph{Heliophila} species was
obtained from the authors of Mandáková et al. (2012). Individual ITS I
and II sequences for \emph{Aethionema grandiflorum, Alliaria petiolata,
Cardamine matthioli, Chamira circaeoides}, and \emph{Rorippa amphibia}
were downloaded from Genbank.

@@

Out group ( \emph{Aethionema grandiflorum, Alliaria petiolata, Cardamine
matthioli, Chamira circaeoides}, and \emph{Rorippa amphibia}) and
ingroup \emph{Heliophila} ITS I and II sequences were aaligned using
MAFFT (Katoh et al., 2002) with strategy G-INS-I, offset value 0.1, and
all other options set as default. The \(GTR + L\) model of nucleotide
substitution was determined to best fit the data based on AIC using
jModelTest2 (Darriba et al., 2012; Guindon and Gascuel, 2003). A maximum
clade credibility tree with branch lengths as relative time was
estimated by summarizing data from six runs of 100,000,000 generations
of Bayesian Markov chain Monte Carlo conducted in BEAST 2 (Bouckaert et
al., 2014). Model selection and phylogenetic analyses were conducted
through the CIPRES Science Gateway (Miller et al., 2010).

@@

To evaluate the hypothesis that annual and perennial life history
strategies reflect adaptations to alternative drought regimes, we tested
the corresponding prediction that the observed distributions of annual
and perennial \emph{Heliophila} species would be significantly
associated with historic drought frequency. First, we compared the
frequency of drought during the winter, spring, summer, and fall between
raw occurrence records of annual and perennial species by t-tests. To
account for variation in the number of occurrence records per species,
we next calculated the mean drought frequency during the winter, spring,
summer and fall for each species. Because shared evolutionary history of
closely related species can lead to spurious associations between traits
and environments (Felsenstein, 1985), we tested for a relationship
between life history strategy and drought frequency while controlling
for phylogeny using phylogenetic logistic regression (Ives and Garland,
2010).

@@

Out of 8670 \emph{Heliophila} GBIF records, 6634 were for species with
reported life history (Mummenhoff et al., 2005), 2856 had geospatial
data, 2833 did not have geospatial issues, 2684 were located on pixels
classified as land having drought measurements, 2543 were located in
Africa, 2192 were not duplicated. Thus, after all filtering steps, 2192
records for 42 species (Figure \ref{fig:phylogeny}, Table
\ref{tab:speciesmeanstable}) passed for further analyses. The number of
samples varied between species, with a mean of 52.19 samples per
species. Heliophila rigidiuscula had the most records, 201, and
Heliophila cornellsbergia the fewest, 2 (Table
\ref{tab:speciesmeanstable}).

@@

The frequency of drought varied considerably across the ranges of
\emph{Heliophila} species (Figure \ref{fig:mapsboxplots}c-f). This
heterogeneity is expected, given that this is one of the most
climatically diverse regions of the Earth (Sayre et al., 2013). It is
worth noting the east to west cline in drought frequency observed during
the summer, which distinguishes the high drought frequency of the
Kalahari Sands and Namid Desert phytogeographic regions from the low
drought frequency of the Drakensberg Mountains and Coastal Zambesian
phytogeographic regions. In the Cape phytogeographic region there was
finer scale heterogeneity in drought frequency during the summer.

@@

We cannot eliminate the possibility that confounding traits or
environmental variables are the causative factors explaining variation
in the distributions of annual and perennial species. Nevertheless,
these results provide quantitative support for the classic prediction
that annual species are found in environemnts that experience more
frequent drought than perennial species. These findings complement
previous reports of qualitative associations between annuality with
enviroments characterized as having increased aridity (Evans et al.,
2005), alternative precipitation defined habitats (Datson et al., 2008;
Morishima et al., 1984), or greater unpredictability (Cruz-Mazo et al.,
2009). However, to our knowledge this is the first study to demonstrate
a significant association between life history and drought in a
phylogenetic context informed by large scale species distribution data
and long term drought measures.

Unfortunately, herbarium collections and their associated data do not
represent systematic or random sampling of a species distribution.
Significant biases in collecting exist, which we have not necessarily
controlled for here, and may have some effect on our findings, such as a
bias toward collecting near roads or near the locations of natural
history collections (Daru et al., 2018). Future research will benefit
from systematic sampling efforts to avoid these noted biases. However,
the ecosystems of southern Africa include several biodiversity hotspots
and are among the most botanically well sampled regions on Earth (Daru
et al., 2018), suggesting that this may currently be the optimal region
for our analyses of life history distribution. Indeed, we were able to
use 2192 occurrence records to study 42 species, which represents a
significant advance over relying on personal observations to
characterize species distributions.

These findings support classical theoretical predictions about the
adaptive value of annual and perennial life history strategies. Taken
together, they suggest that in \emph{Heliophila}, annual species are
adapted to environments with increased summer droughts by avoiding these
seasons in a dormant seed phase of their life cycle. They also suggest
that perenniality is adaptive in environemnts where droughts are less
frequent. While most previous work has focused on describing the
evolutionary origins of annuality (Andreasen and Baldwin, 2001; Barrett
et al., 1996; Conti et al., 1999; Friedman and Rubin, 2015; Verboom et
al., 2004) there are at least a few other cases where perenniality
appears to have arisen from an annual ancestor (Bena et al., 1998; Tank
and Olmstead, 2008). And while early theory predicted selection for
annuality when adult morality is high (Stearns, 1992), we also find
evidence that the transition to perenniality could be explained by the
historical frequency of drought, a likely cause of mortality in plants.
The phylogeny reveals several transitions from annual to perennial life
history (Figure \ref{fig:phylogeny}a) and that the distributions of
perennial \emph{Heliophila} extend into regions where drought frequency
is low (Figure \ref{fig:mapsboxplots}b, Figure \ref{fig:speciesmaps}).
Perennials may be able to out complete annual relatives in environments
where the infrequency of drought favors strategies that allow plants to
benefit from growth over many seasons. It may also indicate that annuals
rely on drought as a source of disturbance for seedling recruitment when
competing with perennials (Corbin and D'Antonio, 2004). Indeed, no
annual species were observed in the low drought regions of eastern South
Africa (Figure \ref{fig:mapsboxplots}, Figure \ref{fig:speciesmaps}).

These findings suggest that species with locally adaptive life history
strategies could be threatened by rapidly changing drought regimens
(Dai, 2011). This could have impacts on ecosystem functioning and
processes such as carbon cycling if life history traits evolve or the
composition of annual and perennial species changes in response (Garnier
et al., 1997; J. Monroe, Markman, et al., 2018; Roumet et al., 2006).
Furthermore, the frequency of drought may be an important factor when
considering the use of perennial cropping systems (Lelièvre and Volaire,
2009; Parry et al., 2005).

@@

\hypertarget{refs}{}
\leavevmode\hypertarget{ref-aghakouchak2015remote}{}%
AghaKouchak A, Farahmand A, Melton F, Teixeira J, Anderson M, Wardlow
BD, Hain C. 2015. Remote sensing of drought: Progress, challenges and
opportunities. \emph{Reviews of Geophysics} \textbf{53}:452--480.

\leavevmode\hypertarget{ref-R-geiger_a}{}%
Alfaro M, Santini F, Brock C, Alamillo H, Dornburg A, Rabosky D,
Carnevale G, Harmon L. 2009. Nine exceptional radiations plus high
turnover explain species diversity in jawed vertebrates.
\emph{Proceedings of the National Academy of Sciences of the United
States of America} \textbf{106}:13410--13414.

\leavevmode\hypertarget{ref-andreasen2001unequal}{}%
Andreasen K, Baldwin BG. 2001. Unequal evolutionary rates between annual
and perennial lineages of checker mallows (sidalcea, malvaceae):
Evidence from 18S--26S rDNA internal and external transcribed spacers.
\emph{Molecular Biology and Evolution} \textbf{18}:936--944.

\leavevmode\hypertarget{ref-appel1997generic}{}%
Appel O, Al-Shehbaz IA. 1997. Generic limits and taxonomy of hornungia,
pritzelago, and hymenolobus (brassicaceae). \emph{Novon} 338--340.

\leavevmode\hypertarget{ref-R-papaja}{}%
Aust F, Barth M. 2018. papaja: Create APA manuscripts with R Markdown.

\leavevmode\hypertarget{ref-barrett1996comparative}{}%
Barrett SCH, Harder LD, Worley AC. 1996. The comparative biology of
pollination and mating in flowering plants. \emph{Phil Trans R Soc Lond
B} \textbf{351}:1271--1280.

\leavevmode\hypertarget{ref-R-Matrix}{}%
Bates D, Maechler M. 2018. Matrix: Sparse and dense matrix classes and
methods.

\leavevmode\hypertarget{ref-bena1998molecular}{}%
Bena G, Lejeune B, Prosperi J-M, Olivieri I. 1998. Molecular
phylogenetic approach for studying life-history evolution: The ambiguous
example of the genus medicago l. \emph{Proceedings of the Royal Society
of London B: Biological Sciences} \textbf{265}:1141--1151.

\leavevmode\hypertarget{ref-bouckaert2014beast}{}%
Bouckaert R, Heled J, Kühnert D, Vaughan T, Wu C-H, Xie D, Suchard M,
Rambaut A, Drummond A. 2014. BEAST 2: A software platform for bayesian
evolutionary analysis. \emph{PLoS Computational Biology}
\textbf{10}:doi:10.1371/journal.pcbi.1003537.

\leavevmode\hypertarget{ref-brown1996oldlist}{}%
Brown PM. 1996. OLDLIST: A database of maximum tree ages. \emph{Tree
rings, environment, and humanity Radiocarbon} \textbf{1996}:727--731.

\leavevmode\hypertarget{ref-charnov1973life}{}%
Charnov EL, Schaffer WM. 1973. Life-history consequences of natural
selection: Cole's result revisited. \emph{The American Naturalist}
\textbf{107}:791--793.

\leavevmode\hypertarget{ref-cole1954population}{}%
Cole LC. 1954. The population consequences of life history phenomena.
\emph{The Quarterly Review of Biology} \textbf{29}:103--137.

\leavevmode\hypertarget{ref-conti1999phylogenetic}{}%
Conti E, Soltis DE, Hardig TM, Schneider J. 1999. Phylogenetic
relationships of the silver saxifrages (saxifraga, sect. Ligulatae
haworth): Implications for the evolution of substrate specificity, life
histories, and biogeography. \emph{Molecular Phylogenetics and
Evolution} \textbf{13}:536--555.

>>>>

\abstract{
Explaining variation in life history strategies is a long-standing goal
of evolutionary biology. For plants, annual and perennial life histories
are thought to reflect adaptation to environments that differ in the
frequency of stress events such as drought. Here we test this hypothesis
in \textit{Heliophila} (Brassicaceae), a diverse genus of flowering
plants native to Africa, by integrating 34 years of satellite-based
drought measurements with 2192 herbaria occurrence records. Consistent
with predictions from classic life history theory, we find that
perennial \textit{Heliophila} species occur in environments where
droughts are significantly less frequent compared to annuals. These
associations are predictive while controlling for phylogeny, lending
support to the hypothesis that drought related natural selection has
influenced the distributions of these strategies. Additionally, the
collection dates of annual and perennial species indicate that annuals
escape drought prone seasons during the seed phase of their life cycle.
Together, these finding provide empirical support for classic hypotheses
about the drivers of life history strategy in plants - that perennials
out compete annuals in environments with less frequent drought and that
annuals are adapted to environments with more frequent drought by
escaping drought prone seasons as seeds. 
}

@@

Understanding the causes and consequences of life history variation is a
longstanding goal of ecology and evolutionary biology (Cole, 1954). In
plants, life histories are especially diverse, with herbaceous species
that complete their life cycle in a number of weeks to trees that live
for thousands of years (Brown, 1996). Along this continuum an important
division exists, distinguishing annuals which complete their seed to
seed life cycle within a single calendar year from perennials which can
persist over multiple years. Annual plants flower once, set seed,
senesce, and then die, spending at least some portion of the year as a
seed, where they are relatively protected from environmental stress. In
contrast, perennial plants can continue vegetative growth after
reproduction and must survive conditions experienced during all seasons.
These represent fundamentally different life history strategies, but the
ecological factors that explain their evolution and distributions remain
empirically uresolved (Friedman \& Rubin, 2015).\\
Classical theory predict shorter life spans in environments where adult
mortality is high (Charnov \& Schaffer, 1973; Stearns, 1992; Franco \&
Silvertown, 1996). In plants, this has been extended to the hypothesis
that annuality is adaptive when it allows plants to escape drought
(Schaffer \& Gadgil, 1975). Lack of water is perhaps the greatest threat
to survival during vegetative or reproductive growth and annuals can
remain dormant (and protected as a seed) during drought. Thus,
environments with greater seasonal drought frequency may select for
annual life histories that complete reproduction prior to drought prone
seasons. Conversely, environments with less frequent drought may select
for perennial species, which may benefit from multiple bouts of
reproduction and competitive advantage by preventing recruitment of
annual species (Corbin \& D'Antonio, 2004). These predictions have been
supported by the observation of annuals in arid environments in
\emph{Oryza perennis} (Morishima \emph{et al.}, 1984) and
\emph{Oenothera} (Evans \emph{et al.}, 2005). Additionally, annual and
perennial species of \emph{Nemesia} were qualitatively associated with
winter rather and summer rainfall environments respectively (Datson
\emph{et al.}, 2008) and annual species of \emph{Scorzoneroides} were
associated with environments classified as unpredictable (Cruz-Mazo
\emph{et al.}, 2009). However, whether the history frequency of drought
events indeed predicts the distributions annual or perennial life
history strategies has yet to be tested.\\
Here we combine a long-term global dataset of satellite detected drought
events with metadata from natural history collections to test these
classic hypotheses within the African endemic mustard genus,
\emph{Heliophila} L. (Brassicaceae). If annuality is an adaptive
strategy allowing plants to escape drought prone seasons, then drought
frequency should predict the distribution of life history strategies
across landscapes, and annual species should be more commonly associated
with drought prone regions than perennial species. Furthermore, if
annual species have adapted to escape drought prone seasons,
observations of growing annual species (i.e.~occurring in forms other
than seed) should be rare during drought prone seasons. Phylogenetic
relatedness can influence tests of associations between species' traits
and their environments (Felsenstein, 1985; Barrett \emph{et al.}, 1996),
and therefore we assessed the relationship between life history
distribution and drought frequency in a phylogenetic context.

@@

Remotely sensed data is a powerful tool for characterizing seasonal
patterns in drought because it is less limited in spatial and temporal
scope and resolution than weather stations or field observations
(AghaKouchak \emph{et al.}, 2015). To quantify the frequency of drought
during different seasons across landscapes, we used the remotely sensed
Vegetative Health Index (VHI), which measures landscape scale reductions
in plant cover and temperature conditions characteristic of drought
(Kogan, 2001). Generated from data collected by NOAA AVHRR satellites
since 1981, the VHI combines Normalized Difference Vegetation Index
(NDVI) derived measures of vegetative stress (Vegetative Condition Index
- VCI) with temperature stress indicated by anomalies in thermal spectra
(Temperature Condition Index - TCI). The VHI of year \(y\) during week
\(w\) of \([1,52]\) at pixel \(i\) is derived from the following
equations, where \(n\) is the number of years observed.

@@

Thus, VHI measurements are standardized according to conditions
historically observed at each locations. These measurements have been
validated and generally used for evaluating drought risk and predicting
crop yields in agriculture (e.g., Rojas \emph{et al.}, 2011; Kogan
\emph{et al.}, 2016). But they also present a new tool to study seasonal
patterns in the frequency of drought across environments and to test
hypotheses about the effect of drought on ecological and evolutionary
processes (Kerr \& Ostrovsky, 2003). As such, the VHI has been applied
recently to study drought related ecology of natural species and proven
useful for predicting intraspecific variation in drought tolerance
traits and genes (Mojica \emph{et al.}, 2016; Dittberner \emph{et al.},
2018; Monroe \emph{et al.}, 2018b). Here, we accessed VHI data at
\(16km^2\) resolution from 1981 to 2015
(\url{https://www.star.nesdis.noaa.gov/smcd/emb/vci/VH/vh_ftp.php}) to
characterize the seasonal drought frequencies experienced by annual and
perennial \emph{Heliophila} species.

@@

\emph{Heliophila} is a genus of flowering plants endemic to the southern
portion of Africa including the Cape Floristic and Succulent Karoo
Regions. These are among the most botanically diverse environments on
Earth and the \emph{Heliophila} species occurring there are considered
to make up the most diverse genus of the family Brassicaceae (Mummenhoff
\emph{et al.}, 2005; Mandáková \emph{et al.}, 2012). This genus includes
both perennial and annual species and this change in life history
strategy has likely arisen multiple independent times (Appel \&
Al-Shehbaz, 1997; Mummenhoff \emph{et al.}, 2005). Furthermore, the fine
scale climatic heterogeneity of Southern Africa is ideal for studying
the distribution of traits in relation to environmental parameters
(Sayre \emph{et al.}, 2013). We used life histories reported by
Mummenhoff \emph{et al.} (2005), grouping species with annual or
perennial life histories. Perenniality was defined based any form of
perennial life history (e.g., herbs, shrubs, mixed, etc). Because the
nature of species reported with mixed traits were unknown
(i.e.~plasticity vs.~genetic variation), we classified these species
here as perennial since they can maintain vegetative growth after
reproduction at least to some capacity.

@@

Botanists have collected and maintained over 350 million botanical
specimens worldwide over the past 300 years. Herbarium specimens and
their associated metadata have been used since the 1960s to study
species' geographical distributions (reviewed by Willis \emph{et al.}
(2017) and Lang \emph{et al.} (2018)). And as they become digitized
(Soltis, 2017), these collections have been used to study relationships
between trait distributions, geography, and climate (Davis \emph{et
al.}, 2015; Stropp \emph{et al.}, 2016; Wolf \emph{et al.}, 2016;
Václavı'k \emph{et al.}, 2017). To characterize the distributions of
annual and perennial \emph{Heliophila} species, all records for the
genus \emph{Heliophila} were downloaded from the Global Biodiversity
Information Facility (gbif.org) on July 21, 2018 (GBIF, 2018).

@@

An alignment of ITS I and II sequences for \emph{Heliophila} species was
obtained from the authors of Mandáková \emph{et al.} (2012). Individual
ITS I and II sequences for \emph{Aethionema grandiflorum, Alliaria
petiolata, Cardamine matthioli, Chamira circaeoides}, and \emph{Rorippa
amphibia} were downloaded from Genbank.

@@

Out group ( \emph{Aethionema grandiflorum, Alliaria petiolata, Cardamine
matthioli, Chamira circaeoides}, and \emph{Rorippa amphibia}) and
ingroup \emph{Heliophila} ITS I and II sequences were aaligned using
MAFFT (Katoh \emph{et al.}, 2002) with strategy G-INS-I, offset value
0.1, and all other options set as default. The \(GTR + \Gamma\) model of
nucleotide substitution was determined to best fit the data based on AIC
using jModelTest2 (Guindon \& Gascuel, 2003; Darriba \emph{et al.},
2012). A maximum clade credibility tree with branch lengths as relative
time was estimated by summarizing data from six runs of 100,000,000
generations of Bayesian Markov chain Monte Carlo conducted in BEAST 2
(Bouckaert \emph{et al.}, 2014). Model selection and phylogenetic
analyses were conducted through the CIPRES Science Gateway (Miller
\emph{et al.}, 2010).

@@

To evaluate the hypothesis that annual and perennial life history
strategies reflect adaptations to alternative drought regimes, we tested
the corresponding prediction that the observed distributions of annual
and perennial \emph{Heliophila} species would be significantly
associated with historic drought frequency. First, we compared the
frequency of drought during the winter, spring, summer, and fall between
raw occurrence records of annual and perennial species by t-tests. To
account for variation in the number of occurrence records per species,
we next calculated the mean drought frequency during the winter, spring,
summer and fall for each species. Because shared evolutionary history of
closely related species can lead to spurious associations between traits
and environments (Felsenstein, 1985), we tested for a relationship
between life history strategy and drought frequency while controlling
for phylogeny using phylogenetic logistic regression (Ives \& Garland,
2010).

@@

Out of 8670 \emph{Heliophila} GBIF records, 6634 were for species with
reported life history (Mummenhoff \emph{et al.}, 2005), 2856 had
geospatial data, 2833 did not have geospatial issues, 2684 were located
on pixels classified as land having drought measurements, 2543 were
located in Africa, 2192 were not duplicated. Thus, after all filtering
steps, 2192 records for 42 species (Figure \ref{fig:phylogeny}, Table
\ref{tab:speciesmeanstable}) passed for further analyses. The number of
samples varied between species, with a mean of 52.19 samples per
species. Heliophila rigidiuscula had the most records, 201, and
Heliophila cornellsbergia the fewest, 2 (Table
\ref{tab:speciesmeanstable}).

@@

The frequency of drought varied considerably across the ranges of
\emph{Heliophila} species (Figure \ref{fig:mapsboxplots}c-f). This
heterogeneity is expected, given that this is one of the most
climatically diverse regions of the Earth (Sayre \emph{et al.}, 2013).
It is worth noting the east to west cline in drought frequency observed
during the summer, which distinguishes the high drought frequency of the
Kalahari Sands and Namid Desert phytogeographic regions from the low
drought frequency of the Drakensberg Mountains and Coastal Zambesian
phytogeographic regions. In the Cape phytogeographic region there was
finer scale heterogeneity in drought frequency during the summer.

@@

We cannot eliminate the possibility that confounding traits or
environmental variables are the causative factors explaining variation
in the distributions of annual and perennial species. Nevertheless,
these results provide quantitative support for the classic prediction
that annual species are found in environemnts that experience more
frequent drought than perennial species. These findings complement
previous reports of qualitative associations between annuality with
enviroments characterized as having increased aridity (Evans \emph{et
al.}, 2005), alternative precipitation defined habitats (Morishima
\emph{et al.}, 1984; Datson \emph{et al.}, 2008), or greater
unpredictability (Cruz-Mazo \emph{et al.}, 2009). However, to our
knowledge this is the first study to demonstrate a significant
association between life history and drought in a phylogenetic context
informed by large scale species distribution data and long term drought
measures.

Unfortunately, herbarium collections and their associated data do not
represent systematic or random sampling of a species distribution.
Significant biases in collecting exist, which we have not necessarily
controlled for here, and may have some effect on our findings, such as a
bias toward collecting near roads or near the locations of natural
history collections (Daru \emph{et al.}, 2018). Future research will
benefit from systematic sampling efforts to avoid these noted biases.
However, the ecosystems of southern Africa include several biodiversity
hotspots and are among the most botanically well sampled regions on
Earth (Daru \emph{et al.}, 2018), suggesting that this may currently be
the optimal region for our analyses of life history distribution.
Indeed, we were able to use 2192 occurrence records to study 42 species,
which represents a significant advance over relying on personal
observations to characterize species distributions.

These findings support classical theoretical predictions about the
adaptive value of annual and perennial life history strategies. Taken
together, they suggest that in \emph{Heliophila}, annual species are
adapted to environments with increased summer droughts by avoiding these
seasons in a dormant seed phase of their life cycle. They also suggest
that perenniality is adaptive in environemnts where droughts are less
frequent. While most previous work has focused on describing the
evolutionary origins of annuality (Barrett \emph{et al.}, 1996; Conti
\emph{et al.}, 1999; Andreasen \& Baldwin, 2001; Verboom \emph{et al.},
2004; Friedman \& Rubin, 2015) there are at least a few other cases
where perenniality appears to have arisen from an annual ancestor (Bena
\emph{et al.}, 1998; Tank \& Olmstead, 2008). And while early theory
predicted selection for annuality when adult morality is high (Stearns,
1992), we also find evidence that the transition to perenniality could
be explained by the historical frequency of drought, a likely cause of
mortality in plants. The phylogeny reveals several transitions from
annual to perennial life history (Figure \ref{fig:phylogeny}a) and that
the distributions of perennial \emph{Heliophila} extend into regions
where drought frequency is low (Figure \ref{fig:mapsboxplots}b, Figure
\ref{fig:speciesmaps}). Perennials may be able to out complete annual
relatives in environments where the infrequency of drought favors
strategies that allow plants to benefit from growth over many seasons.
It may also indicate that annuals rely on drought as a source of
disturbance for seedling recruitment when competing with perennials
(Corbin \& D'Antonio, 2004). Indeed, no annual species were observed in
the low drought regions of eastern South Africa (Figure
\ref{fig:mapsboxplots}, Figure \ref{fig:speciesmaps}).

These findings suggest that species with locally adaptive life history
strategies could be threatened by rapidly changing drought regimens
(Dai, 2011). This could have impacts on ecosystem functioning and
processes such as carbon cycling if life history traits evolve or the
composition of annual and perennial species changes in response (Garnier
\emph{et al.}, 1997; Roumet \emph{et al.}, 2006; Monroe \emph{et al.},
2018a). Furthermore, the frequency of drought may be an important factor
when considering the use of perennial cropping systems (Parry \emph{et
al.}, 2005; Lelièvre \& Volaire, 2009).

@@

\hypertarget{refs}{}
\leavevmode\hypertarget{ref-aghakouchak2015remote}{}%
\textbf{\textnormal{AghaKouchak A}, \textnormal{Farahmand A},
\textnormal{Melton F}, \textnormal{Teixeira J}, \textnormal{Anderson M},
\textnormal{Wardlow BD}, \textnormal{Hain C}}. \textbf{2015}. Remote
sensing of drought: Progress, challenges and opportunities.
\emph{Reviews of Geophysics} \textbf{53}: 452--480.

\leavevmode\hypertarget{ref-R-geiger_a}{}%
\textbf{\textnormal{Alfaro M}, \textnormal{Santini F}, \textnormal{Brock
C}, \textnormal{Alamillo H}, \textnormal{Dornburg A},
\textnormal{Rabosky D}, \textnormal{Carnevale G}, \textnormal{Harmon
L}}. \textbf{2009}. Nine exceptional radiations plus high turnover
explain species diversity in jawed vertebrates. \emph{Proceedings of the
National Academy of Sciences of the United States of America}
\textbf{106}: 13410--13414.

\leavevmode\hypertarget{ref-andreasen2001unequal}{}%
\textbf{\textnormal{Andreasen K}, \textnormal{Baldwin BG}}.
\textbf{2001}. Unequal evolutionary rates between annual and perennial
lineages of checker mallows (sidalcea, malvaceae): Evidence from
18S--26S rDNA internal and external transcribed spacers. \emph{Molecular
Biology and Evolution} \textbf{18}: 936--944.

\leavevmode\hypertarget{ref-appel1997generic}{}%
\textbf{\textnormal{Appel O}, \textnormal{Al-Shehbaz IA}}.
\textbf{1997}. Generic limits and taxonomy of hornungia, pritzelago, and
hymenolobus (brassicaceae). \emph{Novon}: 338--340.

\leavevmode\hypertarget{ref-R-papaja}{}%
\textbf{\textnormal{Aust F}, \textnormal{Barth M}}. \textbf{2018}.
\emph{papaja: Create APA manuscripts with R Markdown}.

\leavevmode\hypertarget{ref-barrett1996comparative}{}%
\textbf{\textnormal{Barrett SCH}, \textnormal{Harder LD},
\textnormal{Worley AC}}. \textbf{1996}. The comparative biology of
pollination and mating in flowering plants. \emph{Phil. Trans. R. Soc.
Lond. B} \textbf{351}: 1271--1280.

\leavevmode\hypertarget{ref-R-Matrix}{}%
\textbf{\textnormal{Bates D}, \textnormal{Maechler M}}. \textbf{2018}.
\emph{Matrix: Sparse and dense matrix classes and methods}.

\leavevmode\hypertarget{ref-bena1998molecular}{}%
\textbf{\textnormal{Bena G}, \textnormal{Lejeune B},
\textnormal{Prosperi J-M}, \textnormal{Olivieri I}}. \textbf{1998}.
Molecular phylogenetic approach for studying life-history evolution: The
ambiguous example of the genus medicago l. \emph{Proceedings of the
Royal Society of London B: Biological Sciences} \textbf{265}:
1141--1151.

\leavevmode\hypertarget{ref-bouckaert2014beast}{}%
\textbf{\textnormal{Bouckaert R}, \textnormal{Heled J},
\textnormal{Kühnert D}, \textnormal{Vaughan T}, \textnormal{Wu C-H},
\textnormal{Xie D}, \textnormal{Suchard M}, \textnormal{Rambaut A},
\textnormal{Drummond A}}. \textbf{2014}. BEAST 2: A software platform
for bayesian evolutionary analysis. \emph{PLoS Computational Biology}
\textbf{10}: doi:10.1371/journal.pcbi.1003537.

\leavevmode\hypertarget{ref-brown1996oldlist}{}%
\textbf{\textnormal{Brown PM}}. \textbf{1996}. OLDLIST: A database of
maximum tree ages. \emph{Tree rings, environment, and humanity.
Radiocarbon} \textbf{1996}: 727--731.
